---
title: "Intro to tidycensus data"
---

## Goals

-   Learn how to use `load_acs()` and `load_decennial()` to retrieve census data at varying geography levels
-   Learn the differences between ACS data and decennial census data

## Setup

First, let's load the two packages we need for this exercise, which are tidycensus and tidyverse.

```{r setup}
#| output: false

library(tidyverse)
library(tidycensus)
```

## Exercises

Let's begin our intro to tidycensus with the basics of how to query the data using the `get_decennial()` and `get_acs()` functions, which will be the two primary functions we use throughout this unit.

### Part 1

For the first part of today's exercise, we'll introduce `get_decennial()`.

As you probably recall from the assigned reading (Walker):

> The decennial US Census is intended to be a complete enumeration of the US population to assist with apportionment, which refers to the balanced arrangement of Congressional districts to ensure appropriate representation in the United States House of Representatives. It asks a limited set of questions on race, ethnicity, age, sex, and housing tenure.

#### Exercise 1

The `get_decennial()` function has several arguments we can use to specify what data we want to retrieve. The three main ones we'll start with are:

-   `year`: What year do we want to retrieve data for?
-   `geography`: What geographical unit do we want to return? (states, counties, census tracts, metro areas, etc.)
-   `variables`: A variable (or list of variables) that we want to retrieve.

For this first exercise, we'll use the `P1_001N` variable, which is used in the 2020 decennial dataset to represent the total population of an area.

::: callout-note
There are literally hundreds of variables that are available through the tidycensus functions, so we're starting with an easy one. In a moment, I'll show you how to find a list of all available variables.
:::

```{r}
#| label: "Get 2020 population by state"
#| message: false


```

#### Exercise 2

Now let's use the `variables` argument to select multiple variables. We can employ the `c()` function for this:

```{r}
#| label: "Get 2020 urban/rural population by state, no variable names"
#| message: false


```

#### Exercise 3

Something else we can do to polish our output from the Census Bureau API is to rename our variables inside the `c()` function. In this case, our first variable represents the overall population of whatever geographies we select, while our next two variables represent the number of people living in areas classified as urban versus rural. So let's rename them:

```{r}
#| label: "Get 2020 urban/rural population by state with variable names"
#| message: false


```

Note how this changes the output: instead of including the variable codes in the output, the API overwrites those with whatever variable name you give it.

#### Exercise 4

The `get_decennial()` and `get_acs()` functions have another argument we can use to structure our data a little differently and give us ways to calculate percentages from the raw data, and that's to use the `summary_var` argument.

To demonstrate, let's make our `P001001` variable into our summary variable, and then let's calculate a "percentage" variable using `mutate()`:

```{r}
#| label: "Get 2020 urban/rural population by state with percentages"
#| message: false


```

Great! Now let's apply this knowledge in Part 2 as we explore the different levels of geographic data that tidycensus provides us with.

### Part 2

Let's use the second part of this exercise not only to introduce the `get_acs()` function for retrieving ACS data, but also to demonstrate how you can retrieve data at different geography levels.

### Exercise 5

So far we've been looking at state-level geography, but let's check out what happens when we switch to county-level data.

```{r}
#| label: "Get 2023 ACS Hispanic/Latino populations by county"
#| message: false


```

::: callout-important
Note here one of the key distinctions between the data that returned from the ACS versus that which is returned from the decennial census: instead of columns named `value` and `summary_value`, we get `estimate` and `summary_est`, because these values are computed from survey data.

Also note how our `moe` and `summary_moe` values are empty for this data. 
:::

#### Exercise 6

By supplying additional arguments to our functions, we can also zoom in on particular states, and even particular counties.

```{r}
#| label: "Get 2023 ACS NC Hispanic/Latino populations by county"
#| message: false


```

#### Exercise 7

When our search area is small enough, like a particular state or county, it probably makes sense for us to choose a smaller geographic unit to analyze. So now let's look at some data at the level of the individual census tracts within Mecklenburg County, NC:

```{r}
#| label: "Get 2023 ACS NC Hispanic/Latino populations by census tract"
#| message: false


```

::: callout-important
Now that we have moved our focus to a small enough geographical area (specific state and/or specific county), we can use smaller geographic units... in this case, **census tracts**. The ACS data is collected in small geographic units, which allows us now to see the margin of error (`moe`) variable so we can understand the level of uncertainty in our data.
:::

#### Exercise 8

Another option which is especially useful if you're interested in data from urban areas is to search for "CBSA" geographies, which is "Core-Based Statistical Areas." These CBSAs are defined by the federal Office of Management and Budget (OMB), and contain urban areas and their surrounding communities. For example, the CBSA for the Charlotte area contains Mecklenburg about about a dozen surrounding counties in North and South Carolina. 

```{r}
#| label: "Get 2023 ACS White/Black populations by CBSA"
#| message: false


```

#### Exercise 9

Lastly, if you're interested in exploring data within particular school districts (perhaps because you're researching trends within public school systems), you can also use a few different geographies like:

-   school district (primary)
-   school district (secondary)
-   school district (unified)

```{r}
#| label: "Get 2023 ACS White/Black populations by unified school district"
#| message: false


```

### Part 3

Big question: **How do we know what variable codes to look for?**

U.S. census data is **massive** and byzantine in its structure, so finding the right variable and what it means can be a complicated affair.

Luckily, there are a few methods you can use:

#### load_variables

The most direct way to get variable codes and descriptions into RStudio is by calling the `load_variables()` function, which is part of the tidycensus package. Our next lesson focuses very heavily on this method.

```{r}
#| label: "load_variables() demo"

dhc_vars_2020 <- load_variables(dataset = "dhc", year = 2020)
```

#### census.gov

The census.gov website has an interactive data portal where you can explore their tables. Link: [Census Bureau Tables](https://data.census.gov/table){target="_blank"}

This interface gives you some capability to explore the available tables and their variables.

When searching for variables for the 2020 decennial census, I recommend you stick to the **Demographic and Housing Characteristics (DHC) dataset** because this has the most data. Again, this is a huge amount of data to work with, but the DHC dataset contains the most possibilities for your final project and other work you might want to do with census data.

::: callout-note
**October 2025**: Due to the shutdown of the federal government, this website may be non-functional. I would keep this in mind when trying to find variables.
:::

Another option under census.gov is their [developer portal](https://www.census.gov/data/developers.html){target="_blank"}, which has a lot of links to documentation that can be helpful for finding variables.

Here are their [decennial datasets](https://www.census.gov/data/developers/data-sets/decennial-census.html){target="_blank"}

And here are their [ACS 5-year datsets](https://www.census.gov/data/developers/data-sets/acs-5year.html){target="_blank"}

These dataset pages link to helpful tables you can browse with codes and descriptions of their variables, such as:
- [2020 decennial DHC variables](https://api.census.gov/data/2020/dec/dhc/variables.html){target="_blank"}
- [2023 ACS 5-year variables](https://api.census.gov/data/2023/acs/acs5/variables.html){target="_blank"}

::: callout-caution
These tables are massive, so use your browser's built-in search features to do keyword searches for specific words and terms.
:::

#### Social Explorer

[Social Explorer's data library](https://www.socialexplorer.com/data-library){target="_blank"} has a streamlined interface for interacting with census tables from both the ACS and the decennial census. They don't have the entire 2020 decennial census yet (they seem to be lacking the DHC dataset), but they have really good tools for exploring recent ACS datasets.

The trick with the Social Explorer is that they use their own proprietary variable codes for census data, so when you find a variable you like and want to analyze with tidycensus, you need to drill down into the Social Explorer's data dictionary to find the equivalent census variable code.

------------------------------------------------------------------------

::: callout-important
You've reached the end of the exercise!
:::
