---
title: "Wrangling tidycensus data"
---

# Goals

-   Learn how to use `load_variables()` to explore the different variables that are available in tidycensus
-   Use the knowledge gained in our last class session to query more census data and pipe it through a series of transformations to make it ready for analysis.
-   Recreate the housing bar chart from the CRF 2022 team's [interactive dashboard](https://crf-dashboard.rc.davidson.edu/){target="_blank"}.

# Setup

As per usual, let's load our packages.

```{r setup}
#| output: false

library(tidyverse)
library(tidycensus)
```

# Exercises

## Part 1

In Part 1, let's introduce the `load_variables()` function and how to use it to retrieve a table of yearly variables from the Census Bureau.

When using `load_variables()`, there are two arguments we need to supply:

-   `year`: the year of the census data we want to retrieve
-   `dataset` the specific dataset we want to retrieve (e.g., "acs5" for the 5-year ACS estimates)

### Exercise 1

Census data, whether it comes from the decennial census or the ACS, consists of a number of different files. With the decennial census, you can get thorough demographics and housing data from the `dhc` dataset. In older versions of the decennial census, you can get comprehensive data from the "summary files" `sf1` and `sf2`.

Let's use our `load_variables()` function to retrieve the DHC data for the 2020 census:

```{r}
#| label: "Load 2020 DHC variables"
#| message: false


```

Immediately you can see the scale of the task in front of us: there are almost **9,000 variables** in the 2020 decennial census. Each variable has a code and a label, and it belongs to a "concept" which is used to group similar variables together.

### Exercise 2

Recall from our last class that the ACS collects a lot more fine-grained data than the decennial does. This definitely has implications for the amount of data these different data sources can return to us.

To contrast the amount of variables between decennial and ACS, let's use `load_variables()` again, but this time let's zero in on one of our ACS datasets.

```{r}
#| label: "Load 2023 ACS5 variables"
#| message: false


```

We can see that the ACS dataset contains **over 28,000 different variables** for us to potentially analyze.

This illustrates one of the biggest challenges when working with tidycensus data: finding and choosing the right variables to answer your questions! Discovering these variables can often seem like a weird scavenger hunt.

## Part 2

For this next part, let's examine a different ACS dataset, and let's focus in on a specific research question: **How much of their income are residents of Mecklenburg County, NC, spending on housing costs?**

### Exercise 3

First, we need to find the right variables. Let's create a new data frame named `rent_vars` that collects the variables from the 2020 ACS 5-year estimate.

```{r}
#| label: "Load 2020 ACS5 variables"
#| message: false


```

OK, we've got almost 28,000 variables to choose from. Let's try to narrow our search.

### Exercise 4

Let's try a little bit of grouping and summarizing. Pipe the `rent_vars` data frame to show **distinct** concepts and a count of variables for each concept, searching for any concepts that contain the key phrase "PERCENTAGE OF HOUSEHOLD INCOME" (needs to be in all caps):

```{r}
#| label: "Get percentage of household income variables"


```

::: callout-note
Of the ten variables we have available to us here, which of these makes the most sense for our research question? 

Again, we are asking: **How much of their income are residents of Mecklenburg County, NC, spending on housing costs?**
:::

### Exercise 5

Before we go any futher, let's ask ourselves a question: **How many variables do we need to recreate the CRF bar chart?**

Once we've figured out this answer, let's see if we can retrieve the necessary variables from our `rent_vars` data frame.

```{r}
#| label: "Get household income by gross rent"

v_concept <- "HOUSEHOLD INCOME BY GROSS RENT AS A PERCENTAGE OF HOUSEHOLD INCOME IN THE PAST 12 MONTHS"

rent_vars2 <- rent_vars |>
  filter(
    concept == v_concept &
    !str_detect(label, "Not computed") & 
    !endsWith(label, ":")
  ) |>
  mutate(
    income_level = str_split(label, ":!!", simplify = TRUE)[, 2],
    pct_income   = str_split(label, ":!!", simplify = TRUE)[, 3]
  ) |>
  rename(variable = name) |> 
  mutate(pct_income = str_replace_all(pct_income,"percent","%")) |>
  select(variable, income_level, pct_income, concept)

rent_vars2
```

### Exercise 6

Once we've identified the variables we need, it's easy to load them into a character vector using our assignment operator:

```{r}
#| label: "Build a list of the variables we need"


```

## Part 3

Now that we know what variables to use, let's retrieve the actual ACS data and try to recreate the CRF bar chart.

### Exercise 7

First, let's use `get_acs()` with the `acs_variables` vector and the `v_year` variable to retrieve the data, then let's run it through a pipeline to produce a data frame named `crf_data` that contains only the data we need to make this chart.

```{r}
#| label: "Get ACS data using the variables list"
#| message: false


```

### Exercise 8

Now let's use `ggplot2` to visualize the results from our `crf_data` data frame. 

```{r}
#| label: "Visualize income levels"
#| fig-width: 10


```

::: callout-important
**Compare your result to the dashboard. What are the differences?**
:::

## Part 4

Let's use this final part of the exercise to get this chart closer to the final product created by the CRF team.

To make aligning the colors easier, you can use this prepared color palette, which is created by simply stringing together a series of 6-digit hex codes inside of the `c()` function to create a list.

```{r}
#| label: "Set color palette"

crf_palette <- c(
  "#F8766D", 
  "#00A9FF", 
  "#00BE67", 
  "#CD9600", 
  "#7CAE00", 
  "#00BFC4", 
  "#C77CFF", 
  "#FF61CC"
)
```

### Exercise 9

Resolve the differences between your chart and the CRF chart using:

- `factor()`
- `scale_fill_manual()` (using our_palette)
- `scale_x_continuous()`
- `labs()`

```{r}
#| label: "Reproduce CRF bar chart"
#| fig-width: 10


```

------------------------------------------------------------------------

::: callout-important
You've reached the end of the exercise!
:::
