---
title: Homework 2
subtitle: CSC 110 - Data Science & Society
format:
  html:
    fig-width: 8
    fig-height: 4
---

# Introduction

This homework allows you to practice more `ggplot2` visualization methods, based on Chapter 1 of *R for Data Science (2nd edition)*.

## Guidance

- Add code blocks under each question heading, and label each code block according to the question it corresponds to. This makes it easier to find and debug errors when they occur.

- If a question asks you to provide both code and a written response, write your non-code answers BELOW your code blocks.

- **Pay close attention** to the instructions! This will ensure that the course autograder is able to validate your test cases successfully!

# Assignment

## Setup

Run the setup block below to load the data for this homework. When you're done, use the tools in RStudio to examine the dataset.

This data comes to us from the World Bank, specifically from their World Development Indicators database. We're retrieving two variables from the year 2020, `gdp_per_capita` and `life_expectancy`, which we can use to compare countries and regions across the globe.

```{r setup}
#| label: setup
#| output: false

library(tidyverse)
library(WDI)

worldbank <- WDI(
  country = "all",
  indicator = c(
    'gdp_per_capita' = "NY.GDP.PCAP.KD", 
    'life_expectancy' = "SP.DYN.LE00.IN"
  ),
  start = 2020,
  end = 2020,
  extra = TRUE,
  cache = NULL,
  latest = NULL,
  language = "en"
) |> filter(region != "Aggregates")
```

---

## Question 1

Use the `worldbank` dataset to create a bar chart named `Q1` that shows the count of countries in each region of the world. Add descriptive labels using the `labs()` function, including an appropriate title, caption, X axis, and Y axis. Below your visual, explain in 1-2 sentences why a bar chart is an appropriate geom for the region variable.

```{r}
#| label: Question-1


```



---

## Question 2

Create a histogram named `Q2` that shows the distribution of the life expectancy variable across all countries. Remember to add descriptive labels for title, subtitle, X axis, and Y axis. Experiment with adjusting the sizes of the histogram bins and take note of how that changes the appearance of the histogram.

::: callout-note
Although you only need to map a single variable to your chart axis here, you can still add descriptive labels to both X and Y axes. This is highly recommended so your audience can better understand what the data in the chart represents!
:::

```{r}
#| label: Question-2


```

---

## Question 3

Create a new `Q3` plot by copy-pasting your code from Q2, but this time create a density plot instead of a histogram. In 1-2 sentences, how does the appearance of the density plot differ from that of the histogram? 

```{r}
#| label: Question-3


```



---

## Question 4

Create a boxplot named `Q4` that puts the life expectancy variable on the X axis and the region variable on the Y axis. Add title, caption, and X/Y axis labels. In 1-2 sentences, what insights can you gather from this plot? Also, why did I ask you to put region on the Y axis instead of the X axis?

```{r}
#| label: Question-4


```



---

## Question 5

Create a bar chart named `Q5` that visualizes the relationship between region and income. To do so, put region on the Y axis and income on the fill aesthetic. Add title, caption, and X/Y axis labels. In 1-2 sentences, summarize a few key insights you can glean from this chart.

```{r}
#| label: Question-5


```



---

## Question 6

Create a new `Q6` plot by copy-pasting your code from Q5. This time, modify your bar chart in such a way that each bar visualizes **proportions** instead of counts. Roughly what percentage of countries in the "East Asia & Pacific" region are considered upper-middle income? 

```{r}
#| label: Question-6


```



---

## Question 7

Create a histogram named `Q7` that shows the distribution of our life expectancy variable, but this time **facet** the chart by region. Add descriptive labels where needed. Finally, in 1-2 sentences, describe what advantages (and disadvantages) faceting might have when compared to the previous charts, particularly the boxplot.

```{r}
#| label: Question-7
#| message: false


```



---

## Question 8

Create a scatterplot named `Q8` with GDP per capita on the X axis and life expectancy on the Y axis. Add a trend line visual on top of the scatterplot using the default ("loess") method. What can you potentially infer about the relationship between these two variables based on the distribution of points in the plot and the shape of our trend line? Think carefully!

```{r}
#| label: Question-8
#| warning: false
#| message: false


```



---

## Question 9

Create a new plot named `Q9` by copy-pasting your code from Q8 and use `facet_wrap()` to break down the scatterplot by income. What insights can you gather from this faceted scatterplot that might not have been as evident in the Q8 scatterplot?

::: callout-tip
The `facet_wrap()` function has an optional argument: `scales`. Search the documentation in RStudio to see if you can figure out how to use it to adjust the appearance of your facets to make them easier to interpret!
:::

```{r}
#| label: Question-9
#| warning: false


```



---

## Question 10 (Reflection)

Using Markdown headings, add a new section to your document named "Q10" and answer the following questions:

- What did you learn from this homework? 
- Was the R4DS reading helpful for completing this assignment?
- What was the most difficult part of this assignment for you?

---

# Submission

Once you are finished with the assignment, carefully review your work against this checklist to ensure you have done everything appropriately.

- [ ] Correct author is listed in the YAML header
- [ ] All code is clearly visible on questions that require it
- [ ] All code output is clearly visible (does not apply to the Setup section)
- [ ] All empty or redundant code blocks and visuals are removed
- [ ] Your written responses to questions are written under the appropriate section headers

When ready, upload your QMD file to Gradescope so the autograder can review your work!
